\\textsc{Mathematical Description}

### 2022-01-11 100-math-description/GR-sslcm-math-description.Rmd
Update math document straying symbology
Get rid of the "T" convention, and replace with "G". T doesn't work in the JAGS model code, because it is interpreted as "TRUE".

There are several sources of mortality that occur on the upstream migration between the mouth of the Columbia River and the natal tributaries in the Grande Ronde basin.
Downstream of Bonneville Dam (BON, the first dam encountered in this upstream migration), returning adults are subject to predation mortality (survival denoted by $\phi^{SL}_{y,j}$) from sea lions and harvest mortality from fisheries (harvest rate after sea lion predation denoted by $U_{y,k,o}$).
We compiled the best available estimates of these quantities, but did not fit to them explicitly as part of the observation model (as we do with many other data sources) -- instead we passed them to the model as non-stochastic quantities that were assumed to be known without error.
Once fish reach BON, however, their survival to LGR has been monitored (since 2000; $y=10$) via PIT tags and so we did fit to those data and estimate the survival rate between BON and LGR as multivariate logit-normal random variables around deterministic values ($\dot{\phi}^{R^{b} \rightarrow R^{a}}_{o}$) with covariance matrix $\Sigma_{\phi^{R^{b} \rightarrow R^{a}}}$ (constructed of $\sigma_{\phi^{R^{b} \rightarrow R^{a}}_o}$ and $\rho_{\phi^{R^{b} \rightarrow R^{a}}}$ following eq. \ref{eq:generic-Sigma2by2}).
In years without these PIT tag counts, however, the survival from BON to LGR is confounded with ocean survival, and so we assumed BON to LGR survival was equal to the deterministic expectation in these years for parameter identifiability purposes:
\begin{equation}
\logit\left(\boldsymbol{\phi}^{R^{b} \rightarrow R^{a}}_{y,1:n_o}\right)
\begin{cases}
= \logit\left(\boldsymbol{\dot{\phi}}^{R^{b} \rightarrow R^{a}}_{1:n_o}\right) & \mbox{if $y < 10$}\\
\sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{R^{b} \rightarrow R^{a}}_{1:n_o}\right), \Sigma_{\phi^{R^{b} \rightarrow R^{a}}} \right] & \mbox{if $y \ge 10$}
\end{cases}
\label{eq:phi-Rb-Ra-random}
\end{equation}
Examination of the adult data indicated that hatchery-origin adults returned to Grande Ronde populations in years that could not be attributed to hatchery-origin smolt releases to these populations.
Further, the Minam River population (which has no hatchery supplementation program) has had hatchery adults return in many of the years with adult data.
Since there were non-zero observations of hatchery-origin adults in cases without a process model component to produce them, we required a process model mechanism to populate non-zero expected values for the observation model to remain valid based on the available data.
We term this part of the model the "straying model", and apply it only in years/populations in which there is no other way to explain the presence of hatchery-origin adults.
We denote the total number of strays that enter population $j$ in year $y$ as $T_{y,o,j}$; we forced all $T_{y,o=\mathrm{NOR},j} = 0$ and forced all $T_{y,o=\mathrm{HOR},j} = 0$ in years where hatchery-origin smolt releases to population $j$ could have explained non-zero hatchery-origin adult returns in year $y$ at age $k$.
We denote the age composition of these strays by $p^{T}_{k,j}$ (where $\sum_k^{n_k} p^{T}_{k,j} = 1$) and apportioned $T_{y,o,j}$ to age-specific values by $T_{y,k,o,j}=T_{y,o,j} \cdot p^{T}_{k,j}$.

We can now define the abundance of adults arriving to their natal spawning tributaries by age and origin ($R^{a}_{y,k,o,j}$) as:
\begin{equation}
R^{a}_{y,k,o,j} = R^{b}_{y,k,o,j} \cdot \phi^{SL}_{y,j} \cdot (1 - U_{y,k,o}) \cdot \phi^{R^{b} \rightarrow R^{a}}_{y,o} + \left(T_{y,o,j} \cdot p^{T}_{k,j}\right)
\label{eq:get-Ra}
\end{equation}
Upon arrival in the tributary, populations with a weir and hatchery program (all populations except the Minam River, indexed by $j = \mathrm{MIN}$) have some number of adults removed each year for broodstock and as a means to minimize natural spawning of hatchery-origin fish.
These weir removals (denoted by $B_{y,k,o,j}$) were passed to the model as perfectly known values and $B_{y,k,o,j=\mathrm{MIN}} = 0$.
Thus, the abundance of fish passed upstream of the weir and that would have a chance to spawn each year was (the maximum constraint was used to ensure all $S^{b}_{y,k,o,j} > 0$):
\begin{equation}
S^{b}_{y,k,o,j} = \max\left(R^{a}_{y,k,o,j} - B_{y,k,o,j}, 1\right)
\label{eq:get-Sb}
\end{equation}
As a final step before spawning, these fish on the spawning grounds must survive the challenges of finding adequate spawning habitat and avoid succumbing to stress and mortality prior to spawning.
We term this phenomenon "pre-spawn survival", denoted by $\phi^{S^{b} \rightarrow S^{a}}_{y,j}$, which we assume are multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{S^{b} \rightarrow S^{a}}_{j}$) with covariance matrix $\Sigma_{\phi^{S^{b} \rightarrow S^{a}}}$ (constructed of $\sigma_{\phi^{S^{b} \rightarrow S^{a}}_j}$ and $\rho_{\phi^{S^{b} \rightarrow S^{a}}}$ following eq. \ref{eq:generic-Sigma4by4}):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{S^{b} \rightarrow S^{a}}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{S^{b} \rightarrow S^{a}}_{1:n_j}\right), \Sigma_{\phi^{S^{b} \rightarrow S^{a}}}\right]
\label{eq:phi-Sb-Sa-random}
\end{equation}
Thus, the abundance of total successfully spawning individuals ($S^{a}_{y,k,o,j}$) was:
\begin{equation}
S^{a}_{y,k,o,j} = S^{b}_{y,k,o,j} \cdot \phi^{S^{b} \rightarrow S^{a}}_{y,j}
\label{eq:get-Sa}
\end{equation}
It is this $S^{a}_{y,k,o,j}$ value from which total egg production ($E_{y,j}$) is calculated in eq. \ref{eq:get-E} to complete the life cycle.

>>>>

There are several sources of mortality that occur on the upstream migration between the mouth of the Columbia River and the natal tributaries in the Grande Ronde basin.
Downstream of Bonneville Dam (BON, the first dam encountered in this upstream migration), returning adults are subject to predation mortality (survival denoted by $\phi^{SL}_{y,j}$) from sea lions and harvest mortality from fisheries (harvest rate after sea lion predation denoted by $U_{y,k,o}$).
We compiled the best available estimates of these quantities, but did not fit to them explicitly as part of the observation model (as we do with many other data sources) -- instead we passed them to the model as non-stochastic quantities that were assumed to be known without error.
Once fish reach BON, however, their survival to LGR has been monitored (since 2000; $y=10$) via PIT tags and so we did fit to those data and estimate the survival rate between BON and LGR as multivariate logit-normal random variables around deterministic values ($\dot{\phi}^{R^{b} \rightarrow R^{a}}_{o}$) with covariance matrix $\Sigma_{\phi^{R^{b} \rightarrow R^{a}}}$ (constructed of $\sigma_{\phi^{R^{b} \rightarrow R^{a}}_o}$ and $\rho_{\phi^{R^{b} \rightarrow R^{a}}}$ following eq. \ref{eq:generic-Sigma2by2}).
In years without these PIT tag counts, however, the survival from BON to LGR is confounded with ocean survival, and so we assumed BON to LGR survival was equal to the deterministic expectation in these years for parameter identifiability purposes:
\begin{equation}
\logit\left(\boldsymbol{\phi}^{R^{b} \rightarrow R^{a}}_{y,1:n_o}\right)
\begin{cases}
= \logit\left(\boldsymbol{\dot{\phi}}^{R^{b} \rightarrow R^{a}}_{1:n_o}\right) & \mbox{if $y < 10$}\\
\sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{R^{b} \rightarrow R^{a}}_{1:n_o}\right), \Sigma_{\phi^{R^{b} \rightarrow R^{a}}} \right] & \mbox{if $y \ge 10$}
\end{cases}
\label{eq:phi-Rb-Ra-random}
\end{equation}
Examination of the adult data indicated that hatchery-origin adults returned to Grande Ronde populations in years that could not be attributed to hatchery-origin smolt releases to these populations.
Further, the Minam River population (which has no hatchery supplementation program) has had hatchery adults return in many of the years with adult data.
Since there were non-zero observations of hatchery-origin adults in cases without a process model component to produce them, we required a process model mechanism to populate non-zero expected values for the observation model to remain valid based on the available data.
We term this part of the model the "straying model", and apply it only in years/populations in which there is no other way to explain the presence of hatchery-origin adults.
We denote the total number of strays that enter population $j$ in year $y$ as $G_{y,o,j}$; we forced all $G_{y,o=\mathrm{NOR},j} = 0$ and forced all $G_{y,o=\mathrm{HOR},j} = 0$ in years where hatchery-origin smolt releases to population $j$ could have explained non-zero hatchery-origin adult returns in year $y$ at age $k$.
We denote the age composition of these strays by $p^{G}_{k,j}$ (where $\sum_k^{n_k} p^{G}_{k,j} = 1$) and apportioned $G_{y,o,j}$ to age-specific values by $G_{y,k,o,j}=G_{y,o,j} \cdot p^{G}_{k,j}$.

We can now define the abundance of adults arriving to their natal spawning tributaries by age and origin ($R^{a}_{y,k,o,j}$) as:
\begin{equation}
R^{a}_{y,k,o,j} = R^{b}_{y,k,o,j} \cdot \phi^{SL}_{y,j} \cdot (1 - U_{y,k,o}) \cdot \phi^{R^{b} \rightarrow R^{a}}_{y,o} + \left(G_{y,o,j} \cdot p^{G}_{k,j}\right)
\label{eq:get-Ra}
\end{equation}
Upon arrival in the tributary, populations with a weir and hatchery program (all populations except the Minam River, indexed by $j = \mathrm{MIN}$) have some number of adults removed each year for broodstock and as a means to minimize natural spawning of hatchery-origin fish.
These weir removals (denoted by $B_{y,k,o,j}$) were passed to the model as perfectly known values and $B_{y,k,o,j=\mathrm{MIN}} = 0$.
Thus, the abundance of fish passed upstream of the weir and that would have a chance to spawn each year was (the maximum constraint was used to ensure all $S^{b}_{y,k,o,j} > 0$):
\begin{equation}
S^{b}_{y,k,o,j} = \max\left(R^{a}_{y,k,o,j} - B_{y,k,o,j}, 1\right)
\label{eq:get-Sb}
\end{equation}
As a final step before spawning, these fish on the spawning grounds must survive the challenges of finding adequate spawning habitat and avoid succumbing to stress and mortality prior to spawning.
We term this phenomenon "pre-spawn survival", denoted by $\phi^{S^{b} \rightarrow S^{a}}_{y,j}$, which we assume are multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{S^{b} \rightarrow S^{a}}_{j}$) with covariance matrix $\Sigma_{\phi^{S^{b} \rightarrow S^{a}}}$ (constructed of $\sigma_{\phi^{S^{b} \rightarrow S^{a}}_j}$ and $\rho_{\phi^{S^{b} \rightarrow S^{a}}}$ following eq. \ref{eq:generic-Sigma4by4}):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{S^{b} \rightarrow S^{a}}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{S^{b} \rightarrow S^{a}}_{1:n_j}\right), \Sigma_{\phi^{S^{b} \rightarrow S^{a}}}\right]
\label{eq:phi-Sb-Sa-random}
\end{equation}
Thus, the abundance of total successfully spawning individuals ($S^{a}_{y,k,o,j}$) was:
\begin{equation}
S^{a}_{y,k,o,j} = S^{b}_{y,k,o,j} \cdot \phi^{S^{b} \rightarrow S^{a}}_{y,j}
\label{eq:get-Sa}
\end{equation}
It is this $S^{a}_{y,k,o,j}$ value from which total egg production ($E_{y,j}$) is calculated in eq. \ref{eq:get-E} to complete the life cycle.
